---
title: "The Eggs Roulette"
---

# Initializing the workspace

```{r}
## some utilities for functional programming and plotting
library(tidyverse)

## our initial basket
EGGS <- c(rep(0, 8), rep(1, 4))

## total number of replicates
NUM_REPLICATES <- 50000

## for reproducibility
RNG_SEED <- 1234
```

# Generating the Sample Space

We'll encode a game as a vector of eggs, with order matching the extraction
order. Odd-positioned eggs are smashed by the guest, even-positioned eggs
are smashed by the host.

Some utility functions:
```{r}
num_raw_eggs_guest <- function(game) {
  if(length(game) == 0) {
    return(0)
  } else {
    return(sum(game[seq(1, length(game), by = 2)]))
  }
}

num_raw_eggs_host <- function(game) {
  if(length(game) < 2) {
    return(0)
  } else {
    return(sum(game[seq(2, length(game), by = 2)]))
  }
}

game_over <- function(game) {
  max(num_raw_eggs_guest(game), num_raw_eggs_host(game)) == 2
}

rgame <- function() {
  eggs <- EGGS
  game <- vector('numeric', length = 0)
  while(!game_over(game)) {
    i <- sample.int(length(eggs), size = 1)
    game <- append(game, eggs[i])
    eggs <- eggs[-i]
  }
  game
}
```

Lets test it:
```{r}
rgame()
```

It works! Now lets generate a few games:

```{r}
set.seed(RNG_SEED)
games <- replicate(NUM_REPLICATES, rgame(), simplify = FALSE)
```

```{r}
length(games)
```

```{r}
str(games[1:10])
```

# Convergence

What's the probability of extracting a raw egg at the first try?

By simple arithmetic:
```{r}
4/12
```

Using our simulated data:
```{r}
unlucky_first <- map_dbl(games, 1)
mean(unlucky_first)
```

The Law of Large Numbers (LNN) in action:
```{r}
p_n <- cumsum(unlucky_first) / seq_len(NUM_REPLICATES)
plot(p_n, type = "l")
abline(h = 1/3, lty = 2, col = 'blue')
```

```{r}
plot(p_n, type = "l", xlim = c(0, 10000))
abline(h = 1/3, lty = 2, col = 'blue')
```

# Multiple Random Variables

```{r}
X <- map_dbl(games, num_raw_eggs_guest)
Y <- map_dbl(games, num_raw_eggs_host)
Z <- map_dbl(games, length)
```

## Marginal probabilities

We can see for each player probabilities of:
- loosing
- staying clean (0 raw eggs)
```{r}
prop.table(table(X))
prop.table(table(Y))
```

How long can a game take?
```{r}
prop.table(table(Z))
```

There's about a 9% chance for the guest to smash 2 raw eggs in a row.

## Joint probabilities

What's the probability of 2 raw eggs for guest, with host staying clean?
```{r}
prop.table(table(X, Y))
```

What about the specific event from the show? Two raw eggs in a row for guest,
with host staying clean:
```{r}
prop.table(table(Y, Z))
```

That is the event (Y=0, Z=3), for which we estimate a probability of about 7.3%.

Note: 3-way tables in R
```{r}
prop.table(table(X, Y, Z))
```

# Conditional probability

Conditional on the first 2 extractions as seen in the show (raw, boiled), what's
the probability of a new raw egg for the guest?
By simple arithmetic:
```{r}
3/10
```

Using our simulated data and applying the definition of conditional probability:
```{r}
E_conditioning <- map_dbl(games, function(game) (game[1] == 1) & (game[2] == 0))
E_conditioned <- map_dbl(games, function(game) game[3] == 1)
mean(E_conditioning & E_conditioned) / mean(E_conditioning)
```

Using our simulated data, plus intuition:
```{r}
games_raw_boiled <- Filter(function(game) {
  game[1] == 1 & game[2] == 0
  },
  games)
length(games_raw_boiled)
```

```{r}
mean(map_dbl(games_raw_boiled, function(game) game[3] == 1))
```

# Geometric Distribution

How many shows we'll have to watch before seeing 2 raw eggs in a raw, with
a clean host?

```{r}
I_event <- function(game) identical(game, c(1, 0, 1))
ii <- map_dbl(games, I_event)
```

```{r}
prop.table(table(ii))
```

When did the events of interest occur?
```{r}
wt <- which(ii == 1)
head(wt)
```

Thus the waiting times:
```{r}
Y <- c(wt[1], diff(wt))
py <- prop.table(table(Y))

head(as.data.frame(py))
tail(as.data.frame(py))

plot(py)
```

How does our empirically determined distribution match the theoretical geometric distribution?

```{r}
ty <- as.tibble(py)
ty$Y <- as.numeric(as.character(ty$Y))
ty$Prob <- dgeom(x = ty$Y, prob = 8/110)
ty <- rename(ty, empirical = n, theoretical = Prob)
head(ty)
tail(ty)
```

Lets plot them side by side:
```{r}
d <- gather(ty, key = "method", value = "probability", empirical, theoretical) %>%
  filter(Y <= 50)
ggplot(d, aes(Y, probability, fill = method)) +
  geom_bar(stat = 'identity', position = position_dodge()) +
  theme_bw()
```

Not bad at all!

## Expected Value

Now, what are its expected value and variance?

By Monte Carlo approximation:
```{r}
c(mean = mean(Y), var = var(Y))
```

## CLT approximation

Out of 200 (independent) game shows, what's the probability of observing
more than 10 short games (3 trials, clean host)?

By Monte Carlo approximation:
```{r}
n <- 200
p <- 8/110
set.seed(RNG_SEED)
Y <- rbinom(n = NUM_REPLICATES, size = n, prob = p)
mean(Y > 10)
```

Using the asymptotic Normal approximation:
```{r}
hist(Y, prob = TRUE, col = 'gray', breaks = 30)
curve(dnorm(x, mean = n * p, sd = sqrt(n * p * (1 - p))), add = TRUE, col = 'blue')
```

```{r}
pnorm(10, mean = n * p, sd = sqrt(n * p * (1 - p)), lower.tail = FALSE)
```

And finally, the exact value:
```{r}
pbinom(10, size = n, prob = p, lower.tail = FALSE)
```

What if we considered 2000 trials rather than 200?
```{r}
n <- 2000
p <- 8/110
set.seed(RNG_SEED)
Y <- rbinom(NUM_REPLICATES, size = n, prob = p)
```

```{r}
hist(Y, prob = TRUE, col = 'gray', breaks = 30)
curve(dnorm(x, mean = n * p, sd = sqrt(n * p * (1 - p))), add = TRUE, col = 'blue')
```

```{r}
curve(pbinom(x, size = n, prob = p), from = -2, to = 200)
curve(pnorm(x, mean = n * p, sd = sqrt(n * p * (1 - p))), lty = 2, add = TRUE, col = 'blue')
```

The CDFs are pretty much identical.
